---
title: "Working with collect() promises in reactive contexts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with collect() promises in reactive contexts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`jsplyr` keeps your data in the browser, so `collect()` has to fetch it back
over an asynchronous round-trip. Because of that, **`collect()` returns a
[promise](https://rstudio.github.io/promises/), not a data frame**.

Reactive *outputs* such as `shiny::renderTable()` or `DT::renderDT()` understand
promises and resolve them for you, which is why a plain `collect()` inside a
render function "just works":

```{r}
output$mtcars_tb <- DT::renderDT({
  lazy_mtcars_query() |>
    dplyr::collect()
})
```

Every other reactive context — `reactive()`, `eventReactive()`,
`observeEvent()` and `observe()` — hands you the promise as-is. The value
resolves later, so you cannot use the result of `collect()` synchronously on the
next line; treating it as a data frame directly gives you a promise object
instead of your rows. You handle it with `promises::then()` or the re-exported
`%...>%` pipe.

## Two ways to unwrap a promise

Both of these work in *any* context — `reactive()`, `eventReactive()`,
`observeEvent()` and `observe()` alike. Pick whichever reads better; the next
section covers where each context differs.

### `promises::then()`

`then()` registers a callback that runs once the data arrives. It returns a new
promise, so it composes:

```{r}
lazy_mtcars_query() |>
  dplyr::collect() |>
  promises::then(function(df) {
    # `df` is the collected tibble
    head(df)
  })
```

### `%...>%` pipe

The `promises` package ships a "promise pipe", `%...>%`, that pipes the
*resolved* value into the next expression. It reads just like a regular pipe but
waits for the promise to settle first. `jsplyr` re-exports it, so you do **not**
have to import `promises` yourself:

```{r}
lazy_mtcars_query() |>
  dplyr::collect() %...>%
  head()
```

Use `%...!%` to handle errors from the promise chain:

```{r}
lazy_mtcars_query() |>
  dplyr::collect() %...>%
  head() %...!%
  (function(err) {
    shiny::showNotification(conditionMessage(err), type = "error")
  })
```

## Where to use it

The operators above are the same everywhere. What changes between contexts is
what you do with the resulting promise: a reactive **returns** it, an observer
**consumes** it for its side effects.

### `reactive()` / `eventReactive()` — return the promise

Return the promise (optionally chained) from the reactive. Downstream consumers
receive the promise and resolve it in turn; a render output does this
automatically. Either operator is fine — use `%...>%` (or `then()`) when you
want to transform the data so the reactive still returns a promise:

```{r}
# Return the promise as-is.
collected <- shiny::reactive({
  lazy_mtcars_query() |>
    dplyr::collect()
})

# eventReactive works the same way — it just gates on an event first.
collected_on_click <- shiny::eventReactive(input$go, {
  lazy_mtcars_query() |>
    dplyr::collect()
})

# Transform inside the reactive and still return a promise.
collected_summary <- shiny::reactive({
  lazy_mtcars_query() |>
    dplyr::collect() %...>%
    head(10)
})

# A render output resolves whichever promise it receives.
output$mtcars_tb <- DT::renderDT({
  collected()
})
```

### `observeEvent()` / `observe()` — consume for side effects

Observers have no return value, so resolve the promise and act on the result
inside the callback. Again, either operator works:

```{r}
# With then().
shiny::observeEvent(input$compute, {
  lazy_mtcars_query() |>
    dplyr::collect() |>
    promises::then(function(df) {
      print(head(df))
      shiny::showNotification(paste(nrow(df), "rows collected"))
    })
})

# With the pipe.
shiny::observeEvent(input$compute, {
  lazy_mtcars_query() |>
    dplyr::collect() %...>%
    head() %...>%
    print()
})
```

Add `%...!%` to catch errors from anywhere earlier in the chain — for example a
failed round-trip or an error thrown while transforming the result:

```{r}
shiny::observeEvent(input$compute, {
  lazy_mtcars_query() |>
    dplyr::collect() %...>%
    head() %...>%
    print() %...!%
    (function(err) {
      shiny::showNotification(
        paste("Failed to collect:", conditionMessage(err)),
        type = "error"
      )
    })
})
```
