---
title: "Working with collect() promises in reactive contexts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with collect() promises in reactive contexts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

`jsplyr` keeps your data in the browser, so `collect()` has to fetch it back
over an asynchronous round-trip. Because of that, **`collect()` returns a
[promise](https://rstudio.github.io/promises/), not a data frame**.

Reactive *outputs* such as `shiny::renderTable()` or `DT::renderDT()` understand
promises and resolve them for you, which is why a plain `collect()` inside a
render function "just works":

```{r}
output$mtcars_tb <- DT::renderDT({
  lazy_mtcars_query() |>
    dplyr::collect()
})
```

Every other reactive context — `reactive()`, `eventReactive()`,
`observeEvent()` and `observe()` — hands you the promise as-is. The value
resolves later, so you cannot use the result of `collect()` synchronously on the
next line; treating it as a data frame directly gives you a promise object
instead of your rows. There are two ways to unwrap it: `promises::then()` and
the `%...>%` pipe.

## Unwrapping a promise with `then()`

`then()` registers a callback that runs once the data arrives. It reads like an
ordinary function call, returns a new promise so it composes, and works in
*any* context — `reactive()`, `eventReactive()`, `observeEvent()` and
`observe()` alike:

```{r}
lazy_mtcars_query() |>
  dplyr::collect() |>
  promises::then(function(df) {
    # `df` is the collected tibble
    head(df)
  })
```

Pass `onRejected` to handle errors from the promise chain:

```{r}
lazy_mtcars_query() |>
  dplyr::collect() |>
  promises::then(
    onFulfilled = function(df) head(df),
    onRejected = function(err) {
      shiny::showNotification(conditionMessage(err), type = "error")
    }
  )
```

### Terser alternative: the `%...>%` pipe

If you prefer an operator, the package ships a "promise pipe", `%...>%`, that
pipes the *resolved* value into the next expression. `jsplyr` re-exports it
(along with `%...!%` for errors), so you do **not** have to attach `promises`
yourself. It is equivalent to the `then()` calls above, just more compact:

```{r}
lazy_mtcars_query() |>
  dplyr::collect() %...>%
  head() %...!%
  (function(err) {
    shiny::showNotification(conditionMessage(err), type = "error")
  })
```

## A practical example: updating an input from a computed value

A common task is to set the value of an input — say a `selectInput` — from a
number computed in the browser. Functions like `shiny::updateSelectInput()` are
**side effects**, not reactive outputs. They do not understand promises, so you
cannot pass `collect()`'s result straight to `selected` — you would hand it a
promise object instead of your value. Resolve the promise first and act on the
value once it arrives.

The cleanest approach keeps the `updateSelectInput()` call **outside** the
`jsplyr` pipeline. Store the resolved value in a `reactiveVal` and let a
separate observer update the input. This decouples "compute the value" from
"update the input":

```{r}
# Holds the value computed in the browser.
oldest_age <- shiny::reactiveVal(NULL)

# Pipeline: compute max(age) and store it. No UI update here.
shiny::observeEvent(input$update, {
  lazy_data() |>
    dplyr::summarise("max_age = max(age)") |>
    dplyr::collect() |>
    promises::then(function(df) {
      oldest_age(df$max_age)
    })
})

# Separate observer: update the input when the computed value changes.
shiny::observeEvent(oldest_age(), {
  shiny::updateSelectInput(
    session,
    inputId = "age",
    selected = oldest_age()
  )
})
```

The same pipeline with the `%...>%` pipe:

```{r}
shiny::observeEvent(input$update, {
  lazy_data() |>
    dplyr::summarise("max_age = max(age)") |>
    dplyr::collect() %...>%
    (function(df) oldest_age(df$max_age))
})
```

`then()` waits for the collected value to arrive, then its callback receives the
result tibble so `df$max_age` is written into the `reactiveVal`. The second
observer reacts to that change and performs the update.

If you prefer to keep the computation in a reactive expression instead of an
observer, return the collected promise from an `eventReactive()` (gated on the
button) and resolve it in a separate `observe()`:

```{r}
# Gated on the button: returns the collect() promise. Computes only.
oldest_age <- shiny::eventReactive(input$update, {
  lazy_data() |>
    dplyr::summarise("max_age = max(age)") |>
    dplyr::collect()
})

# Separate observer: resolve the promise and update the input.
shiny::observe({
  oldest_age() |>
    promises::then(function(df) {
      shiny::updateSelectInput(
        session,
        inputId = "age",
        selected = df$max_age
      )
    })
})
```

The same observer with the `%...>%` pipe:

```{r}
shiny::observe({
  oldest_age() %...>%
    (function(df) {
      shiny::updateSelectInput(
        session,
        inputId = "age",
        selected = df$max_age
      )
    })
})
```

`eventReactive()` returns the promise as-is, so `oldest_age()` is a promise;
the `observe()` resolves it with `then()` and updates the input. A plain
`reactive()` works the same way if you want the value recomputed whenever its
dependencies change rather than only on a button press.

See `inst/example_apps/app_update_select.R` for a complete runnable app showing
both approaches.
